---
title: "Single Cell RNASeq analysis walkthrough"
author: "Eric J. Kort"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: true
    css: custom.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Code conventions

Most of the code below is meant to be entered at the R console, and is 
annotated like this:

```{r}

# Code formatted like this should be pasted into an R session
print("hello world")

```

A few code snippets will need to be entered at the command line (or "Terminal").
Command line code is annotated like this:

```{bash}

# this is code to be entered at the command line
echo Hello world

```

# Prerequisites {#prerequisites}

This document provides a guided analysis of a scRNASeq dataset. The only 
pre-requisites are access to a "well formed" computer/server with R installed. 
By "well formed" I mean that the computer you are working on will need the 
usual GNU toolchain and related tools (gcc, gfortran, libssl, hdf5 libraries, 
etc). MacOS uses clang and that toolchain is fine too. 

## Ubuntu and other Linux distros

From a fresh installation of Ubuntu 18.04, the following will ensure the 
requisite tools are installed. For other flavors of Linux, the steps should 
be analogous, though not necessarily exactly the same depending on which 
packaging system you are using.

```{bash}
# switch to root shell
sudo -s

# install build tools
apt update
apt install -y build-essential
apt intall -y gfortran

# The version of R in the ubuntu releases always lags, so we need to add 
# the R-project repository, get its security key, and then install R

# a little shell magic to get our Ubuntu release's "codename"
RELEASE=$(lsb_release -as 2>/dev/null | tail -n1)

echo "deb https://cloud.r-project.org/bin/linux/ubuntu ${RELEASE}-cran40/" >> /etc/apt/sources.list
apt-key adv --keyserver keyserver.ubuntu.com --recv-keys 51716619E084DAB9
apt  update
apt -y install r-base

# the packages we will need have some system level dependencies. 
# so we will install those
apt -y install libssl-dev
apt -y install libcurl4-openssl-dev
apt -y install libxml2-dev

```

Verify your R version is >= 4.0.2 as follow:

```{bash}
R --version
```

## MacOS

I do not have access to a factory fresh MacOS installation at the moment, but 
I think the clang environment provided by Apple should include all the 
dependencies you need (please open an issue on github if this claim is false).

First, open a terminal window and see if you have clang installed:

```{bash}
clang -v
```

If that returns a version (like `Apple clang version 11.0.0`), you should be 
all set. If not, open run this command:

```{bash}
command xcode-select --install 
```

If you have not already done so, download and install R and (highly recommended) 
RStudio according to the directions [here](https://rstudio.com/products/rstudio/download/#download).

## Windows

Windows binaries for the necessary tools should be available from CRAN, so 
simply download and install R and (highly recommended) 
RStudio according to the directions [here](https://rstudio.com/products/rstudio/download/#download).

You will not be able to build source-only packages with this setup, so your 
mileage may vary. If necessary, you could set up a build environment on your 
Windows box. The R team very helpfully provides the necessary software (based on 
gcc, mingw, and msys) and instructions for building R and its packages from 
source (see instructions [here](https://cran.r-project.org/bin/windows/Rtools)). 

**Note that you do not need to build R from source**, you just need these tools
installed and configured as explained at the link above there to build packages
from source. In particular, there is currently a conflict between the `cpp11`
package and the `tidyr` package that requires building the latest `cpp11`
version on github from source (see below) in order to install `tidyr` (required
by Seurat and, I assume, Monocle as well).

# Installing analysis tools

## Installing Seurat

We will be using the Seurat and Monocle (version 3) packages for the analyses
here. Seurat is available on CRAN so installation should be easy. Try the 
following. If it fails, read the next section. 

```{r eval=FALSE}
install.packages('Seurat')
```
## If Seurat installation fails

While installing Seurat should be easy, at the moment there is a conflict
between the latest release version of the `cpp11` and `tidyr` (on which Seurat
depends). So we need to do a little work around.to install the development
version of `cpp11` that fixes that. **Note:** This requires building from
source, so you will need to have gcc or clang or mingw/msys (Window) installed.
See additional notes in the [prerequisites section above](#prerequisites).

```{r eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("r-lib/cpp11")
install.packages('Seurat')
```

## Installing Monocle3\

The latest Monocle must be installed from github, and has some pre-requisites 
that must be installed. The pre-requisites are part of the Bioconductor 
project so we will add the Bioconductor repositories to the list of searched 
repositories when installing packages.

```{r eval=FALSE}

setRepositories(ind=1:6)
install.packages(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'Matrix.utils'))

# install leidenbase and monocle3 from github
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

# install_github will error out if installed binary packages were built under 
# version of R newer than what we have locally. Change setting to ignore these 
# warnings.
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")

devtools::install_github('cole-trapnell-lab/leidenbase')
devtools::install_github('cole-trapnell-lab/monocle3')

```

# Obtaining Dataset

For this walkthrough, you will need an example dataset from 10X Genomics. 
You can download that dataset 
(here)[https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz]
Decompress that file after you download it.

From the command line you can obtain the dataset and decompress it as follows:

```{bash}
# if you have wget installed
wget https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz

# if you don't
curl -o pbmc3k_filtered_gene_bc_matrices.tar.gz https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz

tar -zxvf pbmc3k_filtered_gene_bc_matrices.tar.gz
```

# Loading data
Now that we have some data, we need to load it into a data object that can be 
used by Seurat or Monocle.

```{r}
library(Seurat)
library(monocle3)

# Seurat
pbmc.data <- Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)

#monocle
cds <- load_mm_data(mat_path = "filtered_gene_bc_matrices/hg19/matrix.mtx", 
                    feature_anno_path = "filtered_gene_bc_matrices/hg19/genes.tsv", 
                    cell_anno_path = "filtered_gene_bc_matrices/hg19/barcodes.tsv")
colnames(rowData(cds)) <- "gene_short_name"

# this will fail if there are duplicate gene names
rownames(cds) <- rowData(cds)$gene_short_name

# filter as above
rc <- apply(counts(cds), 1, function(x) { sum(x > 0) })
cc <- apply(counts(cds), 2, function(x) { sum(x > 0) })
cds <- cds[which(rc >= 3), which(cc >= 200)]
```

You can also convert between the two formats.

```{r}

seuratToMonocle3 <- function(x, assay = NULL) {
  # based on as.CellDataSet function in Seurat, updated for Monocle3
  if(is.null(assay))
    assay <- DefaultAssay(x)

  counts <- GetAssayData(object = x, assay = assay, slot = "counts")
  cell_metadata <- x[[]]
  feature_metadata <- x[[assay]][[]]
  if (!"gene_short_name" %in% colnames(x = feature_metadata)) {
    feature_metadata$gene_short_name <- rownames(feature_metadata)
  }
  new_cell_data_set(counts,
                         cell_metadata = cell_metadata,
                         gene_metadata = feature_metadata)
}

cds <- seuratToMonocle3(pbmc)

# going the other way is simpler

pbmc2 <- CreateSeuratObject(counts = counts(cds), project = "pbmc3k", min.cells = 3, min.features = 200)
```

# QC And Preprocessing

We first need to normalize the data by adjusting for differences in sequencing 
depth and log transforming (after multiplying by 10,000)

```{r}

pbmc <- NormalizeData(pbmc)

# for Monocle, normalization and dimension reduction occur in a single step.
# So this step will take a bit.
cds <- preprocess_cds(cds, num_dim = 100)

```

In the above, Monocle scaled and performed PCA along with normalization. We need 
to perform those steps separately in Seurat.

```{r}

pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
